A content-sensitive framework for modeling episodic memory reveals event-like structure in naturalistic experience, recall, and neural processing

### 2019-08-29 paper/main.tex
updated stats in text, moved topics fig

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modelling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 video timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries detected by the HMM are denoted in yellow (36 events detected).  \textbf{C.} Average topic vectors for each of the 36 video events. \textbf{D.} Topic vectors for each of 294 sentences spoken by an example participant while recalling the video.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (37 events detected).  \textbf{F.} Average topic vectors for each of the 37 recalled events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of video events (Panel C) and recalled events (from the example participant; Panel F).  For similar plots for all participants see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of video events and recalled events (across all 17 participants).  To create the figure, each recalled event was assigned to the video event with the most correlated topic vector (yellow boxes in panels G and H).  The heat maps in each panel were created using \texttt{Seaborn}~\citep{WaskEtal16}.}
\label{fig:model}
\end{figure}

@@

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean: 0.860, SEM: 0.016) and that participants with higher temporal clustering scores also tended to recall more events (Pearson's $r(15) = 0.50, p = 0.042$).

% ------ NOTE: ------
% Semantic clustering correlation with number of events no longer significant.  Andy referenced possible bug in code in slack message a while ago, need to make sure this is using his fixed version
% - Paxton
%Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean: 0.888, SEM: 0.014) and participants who exhibited stronger semantic clustering remembered more video events overall (Pearson's $r(15) = 0.55, p = 0.02$).

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our framework produces a content-based model of individual stimulus and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel metrics for quantifying naturalistic memory representations:~\textit{precision} and \textit{distinctiveness}.  We define precision as the average correlation between the topic proportions of each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, a higher precision score is correlated to both hand-annotated memory performance (Pearson's $r(15) = 0.62, p = 0.08$) and the number of recall events estimated by our model (Pearson's $r(15) = 0.64, p = 0.006$). A second novel metric we introduce here is distinctiveness, or how unique the recall description was to each video event. We define distinctiveness as 1 minus the average of all non-matching recall events from the video-recall correlation matrix. We hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  Similarly to precision, we find that the more distinct participants recalls are (on average), the more they remembered (hand-annotated memory: Pearson's $r(15) = 0.83, p < 0.001$; model-derived memory: Pearson's $r(15) = 0.71, p = 0.001$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.
% ------ NOTE: ------
% Correlations between precision/distinctiveness and number of events need to be updated once analysis is re-run
% - Paxton

@@

\bibliography{memlab}

>>>>

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modelling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 video timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries detected by the HMM are denoted in yellow (36 events detected).  \textbf{C.} Average topic vectors for each of the 36 video events. \textbf{D.} Topic vectors for each of 294 sentences spoken by an example participant while recalling the video, resampled to match the timseries of the video.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (37 events detected).  \textbf{F.} Average topic vectors for each of the 37 recalled events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of video events (Panel C) and recalled events (from the example participant; Panel F).  For similar plots for all participants see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of video events and recalled events (across all 17 participants).  To create the figure, each recalled event was assigned to the video event with the most correlated topic vector (yellow boxes in panels G and H).  The heat maps in each panel were created using \texttt{Seaborn}~\citep{WaskEtal16}.}
\label{fig:model}
\end{figure}

@@

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  Overall, we found that participants heavily clustered video events in their recalls by both temporal proximity (mean: 0.860, SEM: 0.016) and semantic content (mean: 0.888, SEM: 0.014).

% Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean: 0.860, SEM: 0.016) and that participants with higher temporal clustering scores also tended to recall more events (Pearson's $r(15) = 0.50, p = 0.042$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean: 0.888, SEM: 0.014) and participants who exhibited stronger semantic clustering remembered more video events overall (Pearson's $r(15) = 0.55, p = 0.02$).

% ------ NOTE: ------
% Temporal & semantic clustering scores' correlations with number of events & model-estimated memory are no longer significant.  Worth mentioning at all then?
% temporal x hand-annotated: r=.5, p=0.042
% temporal x model-estimated: r=.36, p=.16
% temporal x n events: r=.29, p=.26
% semantic x hand-annotated: r=.32, p=.21
% semantic x model-estimated: r=.33, p=.20
% semantic x n events: r=.48, p=.053
% - Paxton

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our framework produces a content-based model of individual stimulus and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel metrics for quantifying naturalistic memory representations:~\textit{precision} and \textit{distinctiveness}.  We define precision as the average correlation between the topic proportions of each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, a higher precision score is correlated to both hand-annotated memory performance (Pearson's $r(15) = 0.55, p = 0.021$) and the number of recall events estimated by our model (Pearson's $r(15) = 0.66, p = 0.004$). A second novel metric we introduce here is distinctiveness, or how unique the recall description was to each video event. We define distinctiveness as 1 minus the average of all non-matching recall events from the video-recall correlation matrix. We hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  Similarly to precision, we find that the more distinct participants recalls are (on average), the more they remembered (hand-annotated memory: Pearson's $r(15) = 0.62, p = 0.007$; number of events: Pearson's $r(15) = 0.78, p < 0.001$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.
% ------ NOTE: ------
% Correlations between precision/distinctiveness and number of events need to be updated once analysis is re-run
% - Paxton

@@

%\bibliography{memlab}
\bibliography{/Users/paxtonfitzpatrick/Documents/Dartmouth/CDL/CDL-bibliography/memlab.bib}
